% ===========================================================================
% Section 8: Discussion  +  Section 9: Conclusion  (W15 + W16)
%
% Discussion is the designated interpretation section: Robot-mode facts
% live in Sections 5-7; here interpretation is allowed but every claim
% must trace to a table, figure, or the replication archive (footnoted).
%
% Four subsections per tasks.md W15:
%   8.1 Fiscal consolidation as regional policy
%   8.2 MA elasticities in comparative perspective (the Gibbons gap,
%       sharpened by the diagnostics: anchor variants, spatial SEs,
%       theta sensitivity)
%   8.3 Sector-mode specialization (B&P scale economies)
%   8.4 Limitations
%
% Numbers verified against: table CSVs (9, 10, 13, 14), Plan memo /
% diagnostics outputs (theta sweep, refpoint variants, Conley).
%
% CITATION KEYS added this PR: gibbons2024 (STUB), ahlfeldtfeddersen2018.
% ===========================================================================

\section{Discussion}
\label{sec:discussion}

\subsection{Fiscal Consolidation as Regional Policy}
\label{sec:fiscal_policy}

The reform studied here was not designed as regional policy.
Section~\ref{sec:history}
documents that the closures followed fiscal criteria---operating
deficits and traffic density---while the road program followed its own
engineering logic. Yet the combined shock reallocated market access on
a national scale: \maSharePos{}~percent of districts gained market access between
1960 and 1986, because the road expansion's gains outweighed the rail
contraction's losses nearly everywhere
(Figures~\ref{fig:ma_choropleth} and~\ref{fig:ma_counterfactual}). A
policy debate framed as ``closing
unprofitable lines'' was, in its spatial incidence, a large
redistribution of access to markets, with winners and losers that the
fiscal criteria never examined. This is the sense in which fiscal
consolidation acted as regional policy without being one, and it
suggests that the spatial incidence of infrastructure retrenchment
deserves the same scrutiny routinely applied to infrastructure
expansion.

\subsection{Market-Access Elasticities in Comparative Perspective}
\label{sec:comparative}

The closest like-for-like benchmark for our population estimate is
\citet{gibbons2024}, who study the British rail disinvestment of the
same era and estimate a population elasticity of approximately 0.3.
Our full-sample point estimate of \mainPopCoefIVBoth{}
(\mainPopSEIVBoth{}) is below that
benchmark and only marginally distinguishable from zero
($p = \mainPopCoefIVBothP{}$). The
market-access approach itself originates with
\citet{donaldsonhornbeck2016}, whose headline object---agricultural
land values---capitalizes access rather than measuring a population
response, so their estimates inform our method rather than our
magnitudes; output-side responses to access, as in
\citet{ahlfeldtfeddersen2018} for German high-speed rail, are closer
in kind to our sectoral estimates than to the population response.

Three measurement explanations for the low population estimate were
examined and set aside. Re-anchoring market
access at interior points or at each district's largest settlement
does not raise the estimate, and spatial standard errors leave the
headline inference unchanged (both reproduced by the replication
archive's reference-point, urban-anchor, and spatial-standard-error
programs; output files accompany the package). The transshipment
bound, reported in Section~\ref{sec:results_robustness},
reshapes the market-access distribution without changing the
population coefficient's order of magnitude. A candidate substantive explanation, which the research
design cannot test directly, is the bimodal margin itself: the
simultaneous road expansion may have cushioned the population
response to rail-driven access losses. We state it as a hypothesis.

What does move the level is the trade elasticity $\theta$: as
Table~\ref{tab:theta_sweep} in Section~\ref{sec:results_robustness}
shows, the
population coefficient ranges from roughly \sweepBetaThetaOne{} at $\theta = 1$ to
\sweepBetaThetaTwelve{} at $\theta = 12$, and our main value $\theta = \thetaLow{}$ sits in
the interior of that range. The comparison with the literature is
therefore partly a comparison of calibrations, and we treat the
level of the population elasticity as provisional on that choice.
One finding is robust to it: the sectoral contrast
documented below holds at every value of $\theta$ we
examined.\footnote{The sectoral theta sweep is produced by the
replication archive's sectoral-sweep program; its output files
accompany the package. The paper's exhibits report the sweep for the
population outcome only (Table~\ref{tab:theta_sweep}).}

A second difference from the benchmarks is not about calibration:
the spatial granularity of the outcome differs sharply. Our
unit of observation is the district: \nDistricts{} units averaging
\meanDistrictArea~km$^2$, each typically containing several separate
localities --- the 1960 census records approximately 3{,}000
localities nationwide \citep{censo1960pop}, many aligned along the
rail network.
Population reallocation between localities within a district,
including the decline of small rail-side towns, nets out of a
district-level outcome by construction. \citet{gibbons2024} estimate
their elasticity on much finer spatial units, where such
reallocation is part of the measured response. A settled-population
argument --- mobility costs and accumulated local capital damp
responses in mature economies --- applies to the British episode as
much as to ours and therefore cannot explain the gap between the
two estimates; the aggregation margin is consistent with it. Testing
it requires sub-district data, which we leave to future work.

The single-mode character of the benchmarks matters as well.
\citet{gibbons2024} study rail alone; our treatment is the joint
change from a rail contraction and a road expansion measured in one
multimodal index. Structural estimates for the contemporary United
States indicate that the multimodal margin is quantitatively
important: \citet{fuchswong2026} find that ignoring multimodal
flexibility understates the welfare gains from highway improvements
by roughly a fifth in their counterfactuals. In our setting the
analogous statement is descriptive rather than structural: the road
expansion generated market-access gains that more than offset the
rail losses for most districts (Section~\ref{sec:data}), so a
rail-only account of the episode would misstate both the sign and
the geography of the access change.

\subsection{Sector-Mode Specialization}
\label{sec:sector_mode}

The clearest elasticities in this paper are sectoral, not aggregate:
manufacturing production value responds at \mfgValprodCoefIVBoth{}
(\mfgValprodSEIVBoth{}) and manufacturing wage mass at
\mfgWageMassCoefIVBoth{} (\mfgWageMassSEIVBoth{}), while agricultural
outcomes
show no detectable response (Table~\ref{tab:sectoral_iv}). The cost
schedule of \citet{baumgartnerpalazzo1969} offers a candidate
explanation, stated in Section~\ref{ssec:scale_economies} as a
hypothesis: rail is the
cheaper mode above roughly 500 to 1{,}000 tons per day of cargo
density and road below it (Figure~\ref{fig:cost_schedule}). A
modal shift from rail to road therefore lowered transport costs for
the low-density, higher-value shipments characteristic of
manufacturing, while raising them for the bulk flows characteristic
of agriculture. Three exhibits line up behind this reading without
any one of them proving it. The sectoral contrast survives measuring
each sector's market access under its own cost schedule and
strengthens for manufacturing (Table~\ref{tab:sector_matched}); the
aggregate population elasticity rises monotonically as the cost
schedule shifts from rail-favouring to road-favouring
(Table~\ref{tab:density_schedules}); and the counterfactual
decomposition of Section~\ref{sec:counterfactuals} is consistent with
the reading but cannot sharpen it, because the sectoral outcomes are
not observed at the frequency needed to separate modal channels
within a sector. What the alignment does not establish is the
mechanism's exclusivity: the matched-schedule measures strengthen
identification as well as fit (their first stages are stronger), so
part of the sharpening may reflect measurement rather than
economics --- though agriculture's matched schedule also strengthens
its first stage and the agricultural response stays null, so stronger
identification alone does not manufacture a response. Distinguishing
the mechanisms fully would require freight flows by
sector and mode, which do not exist at the district level for this
period.

\subsection{Limitations}
\label{sec:limitations}

Four limitations qualify the results. First, the pre-trends placebo
is not a clean null: post-reform market-access changes predict
1947--1960 population growth at the five-percent level in the OLS
and combined-IV specifications on the
\placeboNIVBoth{}-district placebo subsample
(Table~\ref{tab:pre_trends}), and spatial standard errors do not
soften that rejection. The placebo subsample differs
from the full sample, so selection and pre-trend explanations cannot
be separated with available data; readers should treat the aggregate
population estimates with corresponding caution. Second, the recent
in-migration coefficient in Table~\ref{tab:other_outcomes_iv} is
negative; both a displacement reading and a measurement reading are
consistent with it, and we do not adjudicate. Third, the
hypothetical-road instrument is weak for total market access once
road connectors are re-costed, so identification of the road
component leans on functional form more than we would like; the rail
component, by contrast, is strongly identified throughout. Fourth,
market access is a partial-equilibrium object: it holds destination
populations fixed and abstracts from price responses, so the
estimates are reduced-form gradients, not general-equilibrium
counterfactuals.

\section{Conclusion}
\label{sec:conclusion}

Argentina closed roughly ten thousand
kilometers of railway between 1960 and 1986 while its paved road
network nearly tripled between 1954 and 1986, a rare episode in which
a country
simultaneously contracted one national transport network and expanded
another. We measure the resulting change in access to markets for all
312 districts with a single multimodal index, instrument it with the
Larkin Plan's technical study designations and with hypothetical road
networks, and estimate the response of population and economic
activity on the districts with census coverage.

Three findings summarize the evidence. Aggregate population
reallocation was modest: the full-sample elasticity is small,
significant at the ten-percent but not the five-percent level, and
sensitive to the trade-elasticity calibration. Economic
specialization was not:
manufacturing production and wages responded strongly where market
access rose, while agriculture did not respond, a contrast robust to
every calibration we examined and consistent with the modal cost
structure that made roads cheap for low-density cargo. And the
channel decompositions attribute the population response in
similar parts to rail and road shocks---with the rail component far
better identified---and roughly half of it to infrastructure located
in the district itself.

For policy, the episode illustrates that decisions taken on fiscal
grounds can carry first-order spatial consequences: the geography of
market access was redrawn without any actor being charged with
evaluating that redistribution. For research, the results caution
against transferring market-access elasticities across contexts
without attention to the trade-elasticity calibration, to the
spatial granularity at which outcomes are measured, and to whether
the underlying shock expands access into growing economies or
reallocates it within settled ones.

Four extensions follow directly. A population-weighted urban anchor
built from geocoded census records would settle the reference-point
question definitively. On the sector-mode question, the cost side is
now measured --- each sector's market access under its own cost
schedule (Table~\ref{tab:sector_matched}) --- and the remaining steps
are the demand side, weighting destinations by sectoral economic
mass rather than population, and ultimately freight flows by sector
and mode. Richer within-district data---stations
and depots rather than kilometers---would sharpen the local
decomposition that Section~\ref{sec:mechanisms} begins. And the
digitized multimodal networks and mode-specific cost schedules
assembled here provide the inputs for calibrating a structural model
of multimodal routing with terminal costs
\citep{fuchswong2026} to the episode, which would turn the accounting
decompositions of Section~\ref{sec:counterfactuals} into equilibrium
counterfactuals.
