% ===========================================================================
% Section 3: Data
%
% Per Shapiro Robot-mode writing conventions:
%   - State facts; do not argue. Interpretation belongs in Section 5.
%   - Define every concept before using it.
%   - No forward references to undefined constructs.
%   - Plain, precise language.
%
% All numerical claims in this section trace to either:
%   - code/base/*/clean_*.R                (raw network / census totals)
%   - code/pipeline/03a_build_cost_raster.R (cost-surface parameters)
%   - code/pipeline/04_market_access.R     (MA computation)
%   - data/derived/05_panel/departments_wide_panel.parquet (panel stats)
%
% Numbers marked "FROM PANEL" are pulled directly from the estimation
% panel's manifest; those marked "FROM NETWORK" come from
% rails_by_district.parquet or roads_by_district.parquet.
%
% Eventually these should be replaced with \input{results/scalars.tex}
% AutoFill macros. For now they are inline for readability in the first
% draft.
% ===========================================================================

\section{Data}
\label{sec:data}

The empirical analysis combines three data components. Section~\ref{ssec:networks}
describes the transport network geometries and their changes between
1960 and 1986. Section~\ref{ssec:census} describes the district-level
population and economic outcomes. Section~\ref{ssec:ma} describes how
we construct market access from transport cost surfaces.

\subsection{Transport networks}
\label{ssec:networks}

We combine two georeferenced networks: the railway network at its 1979
status, which we map to 1960 and 1986 via documented closure dates, and
the paved-and-gravel road network at three cross sections (1954, 1970,
1986).

\paragraph{Railways.}
The rail network comes from Argentina's 1979 \emph{Política Ferroviaria}
report \citep{argentina1981politica}, Figure~V--5, scanned and
digitised in QGIS using IGN shapefiles as geographic references. The
resulting shapefile contains 565 MULTILINESTRING segments
in WGS~84, each carrying three attributes that identify its operating
status:
\begin{itemize}
  \item \texttt{status1979} with three values. Value 1 indicates the
  segment was active in both 1979 and 1986 and is therefore counted
  in both the 1960 and 1986 rail stocks. Value 2 indicates the segment
  was closed during the military dictatorship (1976--1983). Value 3
  indicates closure between 1960 and 1966; we count these segments in
  the 1960 stock but not the 1986 stock.
  \item \texttt{studied\_co} with two values. Value 1 indicates the
  segment was one of the 237 rail segments studied in the 1962 Larkin
  Plan, which recommended closures on the basis of freight-density and
  operating-cost criteria. Value 0 indicates the remaining 328 segments
  that the Larkin Plan did not evaluate.
  \item \texttt{recom\_code} with three values (1, 2, 3) encoding the
  Larkin Plan's three operating recommendations for the studied
  segments.
\end{itemize}
We intersect the rail segments with the 312 district polygons from the
IPUMS time-invariant geography (described in Section~\ref{ssec:census}),
compute per-district length in kilometres in EPSG:4326 using
\texttt{sf::st\_length} with s2 geometry, and obtain \texttt{tot\_rails\_1960} and
\texttt{tot\_rails\_1986} for each district. Total rail km falls from 42{,}780
in 1960 to 33{,}303 in 1986, a 22\% reduction. Of the 9{,}477~km lost,
5{,}563~km (59\%) were closed during the dictatorship and 3{,}914~km (41\%)
were closed between 1960 and 1966. Figure~\ref{fig:networks}a shows the
geography of the rail network and highlights the dictatorship closures
in red. 30 of the 312 districts have zero rail km in all periods;
these are concentrated in Patagonia and Tierra del Fuego.

\paragraph{Roads.}
The road network comes from a digitised cross-panel of
Automóvil~Club~Argentino maps for 1954, 1970, and 1986. The resulting
shapefile (\texttt{comparacion\_54\_70\_86.shp}) contains 1{,}741
MULTILINESTRING segments classified into seven \texttt{type2}
categories by their period-specific presence:
\begin{itemize}\setlength\itemsep{0pt}
  \item \texttt{type2}~$\in\{1,5\}$: present in 1954 and 1986 (type 5 is
    absent from the 1970 map and treated as a cartographic error).
  \item \texttt{type2}~$\in\{2,3\}$: added between 1954 and 1986
    (type 2 appears in 1970, type 3 appears in 1986).
  \item \texttt{type2}~$\in\{4,6,7\}$: present only in a subset of
    periods and absent in 1986; these segments are excluded from
    period-total road km.
\end{itemize}
The paved-and-gravel road network (the union of types 1, 2, 3, and 5)
grows from 27{,}513~km in 1954 to 79{,}820~km in 1986, a 190\% increase.
Figure~\ref{fig:networks}b shows the road network and highlights the
segments added between 1954 and 1986 in red.

Three districts, all in Patagonia, lack any segment in the road
shapefile and contribute missing values in the district-level road
variables. 11 districts lost rail without gaining roads and 141 districts
both lost rail and gained roads. The cross-district correlation between
$\Delta$(rail km) and $\Delta$(road km) is $-0.25$.

\paragraph{Hypothetical road networks.}
We also construct four hypothetical road networks that connect Argentina's
largest cities under stylised routing rules. Starting from a node set of
54~cities (all provincial capitals plus all cities with population
$\geq 15{,}000$ in 1960), we compute:
\begin{itemize}\setlength\itemsep{0pt}
  \item \texttt{euc}: all $\binom{54}{2}=1{,}431$ straight-line
    (Euclidean) segments between city pairs.
  \item \texttt{euc\_mst}: the Euclidean minimum spanning tree
    (53 edges).
  \item \texttt{lcp}: all bilateral least-cost paths between city
    pairs, routed on a Faber-2014 construction-cost raster that
    penalises slope, wetland, urban areas, and water crossings.
  \item \texttt{lcp\_mst}: the minimum spanning tree of the LCP
    weights.
\end{itemize}
These networks serve as instruments for the actual road network in
Section~\ref{sec:identification}.

\subsection{Census data and outcomes}
\label{ssec:census}

\paragraph{Geographic unit.}
We work at the district level using IPUMS International's
time-invariant \texttt{geolev2} geography for Argentina, covering 312
mainland districts including Capital Federal and the two Tierra~del~Fuego
jurisdictions. The same polygons are used across the five IPUMS sample
years (1970, 1980, 1991, 2001, 2010), the 1960 and 1947 census
publications, and the agricultural and industrial censuses. The
\texttt{geolev2} identifier is stored as a character string throughout
the analysis to avoid type-coercion bugs on leading zeros.

\paragraph{Population.}
Population in 1960 comes from the digitised \emph{IV Censo Nacional de
Población 1960} tables, reconciled to the IPUMS district geography
through a crosswalk that handles pre-1960 splits and name changes. The
mean 1960 district population is 43{,}856 with a national total of
13.6~million. Population in 1991 comes from the IPUMS microdata
aggregated to \texttt{geolev2}; the mean 1991 district population is
114{,}723 with a national total of 35.8~million. Log population change
$\Delta\log(\mathrm{pop}_{91/60})$ has mean $+1.02$ and standard deviation
$0.53$.

\paragraph{Urban population.}
Urban population shares at the district level come from the 1960
census (classification per INDEC) and from the IPUMS 1991 microdata.
Urban population is not coded consistently in the 1970 and 2010 IPUMS
microdata for Argentina; those years are recorded as missing in the
panel and are not used for urban-outcome regressions. 1960 and 1991 give
us the post-reform window with consistent urban/rural classification.

\paragraph{Pre-reform placebo.}
We supplement the baseline with the 1947 national census, which covers
240 of the 312 districts. A log population change
$\Delta\log(\mathrm{pop}_{60/47})$ for this subsample serves as a placebo
outcome: it measures population growth before any of the 1960--1986
network changes, and should be uncorrelated with the instruments if the
instruments isolate post-1960 network-induced variation.

\paragraph{Industrial and agricultural outcomes.}
Four additional data components carry sectoral outcomes. The 1954 and
1985 industrial censuses (\emph{Censo Industrial} and \emph{Censo
Económico}) provide district-level establishment counts, salaries, and
production values. Only \texttt{nestab} (establishments), \texttt{massal}
(wages), and \texttt{valprod} (production value) are present in both
census years; \texttt{nemp} (workers, 1954) and \texttt{npers} (persons,
1985) are defined differently and are not directly comparable. The 1960
and 1988 agricultural censuses provide \texttt{nexp} (farm operations)
and \texttt{areatot\_ha} (total agricultural area). All monetary
variables are nominal pesos of their census year; real comparisons
across years require a deflator that we defer to robustness.

\paragraph{Geographic controls.}
District-level controls come from external sources. Area in~km$^2$ is
computed from the IPUMS polygon. Mean elevation comes from GTOPO30
\citep{usgs1996gtopo30}. Terrain ruggedness uses the Nunn--Puga TRI
raster \citep{nunn2012ruggedness} derived from GTOPO30. Wheat potential
yield comes from FAO--GAEZ~3.0 \citep{faoiiasa2012gaez}. Caloric
potential (pre-~and post-1500) comes from \citet{galorozak2016}.
Distance to Buenos Aires is the great-circle distance from the centroid
of the Gran~Buenos~Aires polygon in the IGN settlements shapefile.
All continuous controls are standardised (zero mean, unit variance) for
use in the regressions.

\subsection{Market access}
\label{ssec:ma}

\paragraph{Definition.}
Following \citet{donaldsonhornbeck2016},
we define district $i$'s market access in period $t$ for sector
$s$ under elasticity $\theta$ as:
\begin{equation}
  \mathrm{MA}_{i,t,s,\theta} = \sum_{j \neq i}
     \frac{\mathrm{pop}_{j,1960}}{(\tau_{ij,t,s})^{\theta}},
  \label{eq:ma}
\end{equation}
where $\tau_{ij,t,s}$ is the origin--destination transport cost between
the centroids of districts~$i$ and~$j$ under network configuration~$t$
and sector~$s$. Population weights are held fixed at their 1960 values
across all $t$ so that variation in~$\mathrm{MA}_{i,t,s,\theta}$ comes
entirely from variation in transport costs rather than from endogenous
population responses to the network changes.

\paragraph{Cost surface.}
Transport costs $\tau_{ij,t,s}$ are the least-cost-path distances on
a $2{,}399 \times 3{,}090$ raster covering mainland Argentina in the
ESRI:54034 equal-area projection (approximately 1.2~km per cell). The
per-pixel cost $c_i(t,s)$ is:
\begin{equation*}
c_i(t, s) =
\begin{cases}
\text{cost}_{\mathrm{nav}}(s)                      & \text{if pixel $i$ is navigable water,} \\
\text{cost}_{\mathrm{mininfra}}(s)                 & \text{if pixel $i$ has both rail and road,} \\
\text{cost}_{\mathrm{rail}}(s)                     & \text{if pixel $i$ has rail only,} \\
\text{cost}_{\mathrm{road}}(s)                     & \text{if pixel $i$ has road only,} \\
\text{cost}_{\mathrm{land}} \cdot \mathrm{HMI}_i   & \text{if pixel $i$ is off-network land in Argentina,} \\
\text{NA}                                          & \text{otherwise (hard barrier).} \\
\end{cases}
\end{equation*}
Network layers (rail, road, and navigation) are rasterised from the
respective shapefiles after a 1~km buffer to guarantee at least one
full raster cell per linear segment. The Strait of Magellan is buffered
by 5~km so that at least one raster cell per shore becomes navigable,
which lets Dijkstra cross the strait and keeps Tierra~del~Fuego
connected to the mainland. Off-network pixels outside Argentina are
set to NA, which acts as a hard barrier for the least-cost path
algorithm except where a navigable-water pixel overrides the country
mask (e.g.\ the Paraná river border).

\paragraph{Cost parameters.}
Sector-specific cost parameters come from \citet{baumgartnerpalazzo1969}
Table~II. Three sectors $s \in \{0, 1, 2\}$ correspond to three cargo
density scenarios (medium, high, low); Table~\ref{tab:cost_params}
summarises the values.

\begin{table}[htbp]
\centering
\small
\caption{Transport Costs by Mode and Cargo Density
         (1960 pesos per ton-km)}
\label{tab:cost_params}
\begin{tabular}{lccc}
\toprule
              & Low density  & Medium density & High density \\
              & (100 t/day)  & (500 t/day)    & (1{,}000 t/day) \\
\midrule
Road          & 8.266        & 1.777          & 1.000 \\
Railroad      & 13.490       & 1.874          & 0.465 \\
Navigation    & 0.621        & 0.621          & 0.621 \\
\bottomrule
\end{tabular}

\vspace{0.5em}
{\footnotesize
\begin{minipage}{0.85\textwidth}
\emph{Notes}: Ground-transport costs from
\citet{baumgartnerpalazzo1969}. Navigation costs from
\citet{larkin1962}. Medium-density costs are used in the main
analysis (sector $s = 0$); high- and low-density costs (sectors
$s = 1$ and $s = 2$) are reported in extensions exploring
sector-specific effects.
\end{minipage}
}
\end{table}

In particular, rail is cheaper than road at high cargo density
($\text{cost}_\mathrm{rail}(s{=}1) = 0.465 <
\text{cost}_\mathrm{road}(s{=}1) = 1.000$) but more expensive at low
density ($\text{cost}_\mathrm{rail}(s{=}2) = 13.490 >
\text{cost}_\mathrm{road}(s{=}2) = 8.266$), which is the
scale-economies channel that motivates sector-specific analysis. At
medium density, rail and road costs are nearly identical (1.874 vs.\
1.777), so mode choice matters less for the overall market-access
index. We therefore use the medium-density scenario as the main
specification and report the other two as extensions. The
off-network cost $\text{cost}_{\mathrm{land}}$ is
$17.9 \times \text{cost}_\mathrm{road}(s{=}2) / \overline{\mathrm{HMI}}$,
where $\overline{\mathrm{HMI}} = 0.2018$ is the national mean of the
Özak~\citeyearpar{ozak2018} Human Mobility Index. The factor~17.9 is
the ratio of motor-vehicle speed to walking-equivalent speed, calibrated
on a Buenos~Aires--Rosario comparison. This gives
$\text{cost}_{\mathrm{land}} \approx 733.2$; the per-pixel off-network
cost scales with the local HMI value and is one to two orders of
magnitude above on-network cost.

\paragraph{Least-cost path and elasticity.}
We compute $\tau_{ij,t,s}$ by running Dijkstra's algorithm on an
eight-connected grid transition object (\texttt{gdistance::costDistance})
between the 312 IPUMS centroid points. $\tau_{ii,t,s}=0$ by construction
and $\tau$ is symmetric. Dijkstra is run independently for each
(period, sector) combination; a single end-to-end run of the four
main networks (actual~1960, actual~1986, \texttt{stu} instrument,
\texttt{lcp\_mst} instrument) at three sectors takes approximately
45~minutes on a laptop with four cores.

The elasticity $\theta$ controls how sharply distant destinations are
discounted in the MA sum. We follow the applied-trade literature
(\citealt{donaldsonhornbeck2016}; \citealt{eatonkortum2002}) and report
two values, $\theta_{\mathrm{low}} = 4.55$ and $\theta_{\mathrm{high}} =
8.11$. Main results use $\theta_{\mathrm{low}}$; $\theta_{\mathrm{high}}$
appears in the sensitivity tables. [PLACEHOLDER: confirm citations and
whether to add a literature-review paragraph here per the
\texttt{Plan/tasks.md} pending-decision item.]

\paragraph{Change in log MA.}
The main treatment variable is
$\Delta\log\mathrm{MA}_i \equiv \log\mathrm{MA}_{i,1986,s=0,\theta_{\mathrm{low}}}
  - \log\mathrm{MA}_{i,1960,s=0,\theta_{\mathrm{low}}}$.
Across the 312~districts,
$\Delta\log\mathrm{MA}_i$ has mean $+1.56$, standard deviation $2.48$,
minimum $-7.18$ and maximum $+9.58$; 91\% of districts see a positive
change. Figure~\ref{fig:ma_choropleth} maps $\Delta\log\mathrm{MA}_i$
across districts.

\paragraph{Counterfactual decomposition.}
We additionally construct two counterfactual cost surfaces that freeze
one network mode at its baseline while letting the other change:
\begin{itemize}\setlength\itemsep{0pt}
  \item \emph{only-rail}: 1986 rails combined with 1954 roads.
  $\Delta\log\mathrm{MA}^{\text{only-rail}}_i = \log\mathrm{MA}(\text{only-rail})_i
    - \log\mathrm{MA}_{i,1960}$.
  \item \emph{only-road}: 1960 rails combined with 1986 roads.
  $\Delta\log\mathrm{MA}^{\text{only-road}}_i$ defined analogously.
\end{itemize}
At sector~0 and $\theta_{\mathrm{low}}$, $\Delta\log\mathrm{MA}^{\text{only-rail}}$
has mean $-0.40$ (rail contraction is a small market-access loss) and
$\Delta\log\mathrm{MA}^{\text{only-road}}$ has mean $+1.75$ (road
expansion is a large gain). The sum of the two isolated components
($+1.35$) is slightly below the total
$\Delta\log\mathrm{MA} = +1.56$, reflecting a positive complementarity
between the two network shocks. Figure~\ref{fig:ma_counterfactual}
compares the three maps side by side.

\paragraph{Limitations.}
The cost-surface construction imposes three modelling choices that the
Section~\ref{sec:identification} discussion of identification returns to.
First, mode switches are free: Dijkstra can step from a rail pixel to a
road pixel at any grid boundary, whereas in reality such switches
occur at stations or terminals and incur transshipment costs. Second,
the algorithm selects a single cheapest path, not a probability
distribution over routes; in equilibrium firms use multiple parallel
routes. Third, cost parameters are constant across space and time; we
do not model congestion or regional cost heterogeneity. Each of these
simplifications follows the approach in
\citet{donaldsonhornbeck2016} and is common in the market-access
literature.
