% ===========================================================================
% Section 5: Main Results
%
% Robot-mode (Shapiro Step 3): state facts, do not argue; define concepts
% before use; no forward references to undefined constructs; plain precise
% language.
%
% Subsections drafted here (anchored by Tables 8, 9, 10):
%   5.1 First Stage
%   5.2 Population Effects
%   5.3 Sectoral Activity (renamed from "Employment and Sectoral Effects"
%       per PR #49 -- the paper-skeleton employment outcomes could not
%       be implemented; the industrial and agricultural census activity
%       outcomes are used instead. See table_10_sectoral.R header.)
%
% Subsections deferred (need more analysis before drafting):
%   5.4 Other Outcomes   (pending Table 11: education, migration,
%                        employment status)
%   5.5 Robustness       (pending Table 12: alternative theta, alternative
%                        hypo instruments, subsample stability)
%
% All regression numbers in this section are AutoFill macros from
% results/scalars.tex (paper-wide pass after issue #22). Significance
% WORDS ("five-percent level") remain prose and need a manual check
% when numbers move.
%
% CITATION KEYS: none new beyond Sections 1-4.
% ===========================================================================

\section{Main Results}
\label{sec:results}

\subsection{First Stage}
\label{sec:first_stage}

Table~\ref{tab:first_stage} reports the first-stage regressions
corresponding to equation~\eqref{eq:first_stage}. Column~(1) uses only
the Larkin Plan instrument $\Delta \ln \mathrm{MA}^{\mathrm{LP}}_i$,
column~(2) uses only the hypothetical-road instrument
$\Delta \ln \mathrm{MA}^{\mathrm{hypo}}_i$, and column~(3) uses both
jointly. All columns include baseline log market access, baseline log
population, and the six standardized geographic controls. Standard
errors are heteroskedasticity-robust.

The Larkin Plan instrument has a coefficient of \fsLPCoef{} with a robust
standard error of \fsLPSE{} in column~(1). The corresponding first-stage
$F$ statistic is \firstStageFLP{}. The hypothetical-road instrument has a
coefficient of \fsHypoCoef{} (\fsHypoSE{}) in column~(2), with a first-stage $F$ of
\firstStageFHypo{}. In the joint specification in column~(3), both instruments
retain predictive power: the Larkin Plan coefficient is \fsJointLPCoef{} (\fsJointLPSE{})
and the hypothetical-road coefficient is \fsJointHypoCoef{} (\fsJointHypoSE{}); the joint $F$
is \firstStageFBoth{}.

The Larkin Plan instrument alone exceeds the Stock--Yogo rule-of-thumb
threshold of 10. The hypothetical-road instrument alone does not. The
joint specification exceeds the threshold and is adopted as the main
specification throughout this section. The hypothetical-road-only
column is reported so that the two sources of variation can be
inspected separately.

\subsection{Population Effects}
\label{sec:population}

Table~\ref{tab:population_iv} reports estimates of equation~\eqref{eq:main}
for four population outcomes: the log change in total population, in
urban population, and in rural population between 1960 and 1991, and
the level change in the urban share over the same period. Each
outcome panel reports four columns: OLS in column~(1), instrumental
variables using the Larkin Plan instrument in column~(2), the
hypothetical-road instrument in column~(3), and both instruments in
column~(4).

For total population (Panel~A), the OLS coefficient on $\Delta \ln
\mathrm{MA}^{\mathrm{full}}_i$ is \popOLSCoef{} (\popOLSSE{}), significant at the
five-percent level. The IV-LP estimate is \popIVLPCoef{} (\popIVLPSE{}), the IV-Hypo
estimate is \popIVHCoef{} (\popIVHSE{}), and the combined IV estimate is
\mainPopCoefIVBoth{} (\mainPopSEIVBoth{}). The sample is
\mainPopNIVBoth{} districts: the \nDistricts{} mainland districts less Capital
Federal, which is excluded as an observation
(Section~\ref{sec:data}).

For urban population (Panel~B), the OLS coefficient is \urbOLSCoef{} (\urbOLSSE{}),
significant at the five-percent level. The IV-LP estimate is \urbIVLPCoef{}
(\urbIVLPSE{}), the IV-Hypo estimate is \urbIVHCoef{} (\urbIVHSE{}), and the combined IV
estimate is \urbIVBCoef{} (\urbIVBSE{}). The sample is \urbN{} districts, reflecting
districts with positive urban population in both census years.

For rural population (Panel~C), the OLS coefficient is \rurOLSCoef{} (\rurOLSSE{}).
The IV-LP estimate is \rurIVLPCoef{} (\rurIVLPSE{}), the IV-Hypo estimate is \rurIVHCoef{}
(\rurIVHSE{}), and the combined IV estimate is \rurIVBCoef{} (\rurIVBSE{}). The sample is
\rurN{} districts.

For the urban share (Panel~D), all four estimates are close to zero
and statistically indistinguishable from zero (OLS: \urbshrOLSCoef{} (\urbshrOLSSE{});
IV-Both: $\urbshrIVBCoef{}$ (\urbshrIVBSE{})). A one-unit increase in $\Delta \ln
\mathrm{MA}^{\mathrm{full}}_i$ corresponds to a change in the urban
share of less than one percentage point.

Across the three log-change outcomes, the IV point estimates exceed
the OLS estimates by a factor of approximately two to four. This
pattern is consistent with a downward bias in OLS that would arise if
road-driven market-access gains were directed to districts with
declining pre-period trends --- the compensatory allocation logic of
Section~\ref{sec:identification}. The IV point estimates are not
statistically distinguishable from zero at the five-percent level,
though the total-population combined-IV estimate is significant at
the ten-percent level ($p = \mainPopCoefIVBothP{}$) and the
total-population OLS estimate at the five-percent level. The
IV-Both column reports the narrowest confidence intervals among the
three IV columns.

Two mechanisms can produce an IV--OLS gap of this direction, and the
evidence does not separate them. The first is compensatory selection
(Section~\ref{sec:identification}): if road-driven market-access
gains were directed to districts whose populations would have grown
slowly anyway, gains align with negative outcome shocks and OLS
understates the causal response. The opposing efficiency logic ---
closures concentrated where outcomes were already falling --- biases
OLS upward and so cannot explain a gap of this direction. The placebo
evidence bears on which logic dominated: on the subsample where
pre-period growth is observed, post-reform market-access changes
correlate \emph{positively} with 1947--1960 growth
(Section~\ref{sec:pre_trends}), the efficiency-selection pattern,
which taken alone points to upward, not downward, OLS bias and so
weighs against selection of either sign explaining the observed gap.
The second mechanism is attenuation: market
access is a constructed regressor built from digitised networks and
calibrated cost parameters, and classical measurement error in it
biases OLS toward zero. Instrumenting removes the attenuation only to
the extent that the instruments do not share the construction error
--- a condition that is not innocuous here, since both instruments
are themselves market-access measures built from the same digitised
baseline network and the same cost parameters. With the available
data we do not apportion the gap between the two mechanisms.

\input{../results/tables/table_9_population_iv.tex}

\paragraph{Cargo-density cost schedules.}
The estimates above use the medium-density (500~t/day) cost schedule,
under which rail and road costs per ton-km are nearly identical
(Table~\ref{tab:cost_params}), so the market-access index does not
build in a preference for either mode.
Table~\ref{tab:density_schedules} re-estimates the total-population
specifications under the other two \citet{baumgartnerpalazzo1969}
schedules, switching every market-access object --- treatment, both
instruments, and the baseline log-MA control --- to the row's
schedule. The treatment correlates \densHighCorrBase{} (high density)
and \densLowCorrBase{} (low density) with its baseline version, so the
three rows are related but not interchangeable measures. The
combined-IV estimate rises monotonically as the cost schedule shifts
from rail-favouring to road-favouring: \densHighIVBCoef{}
(\densHighIVBSE{}) at high density, \mainPopCoefIVBoth{}
(\mainPopSEIVBoth{}) at the baseline, and \densLowIVBCoef{}
(\densLowIVBSE{}, $p = \densLowIVBP{}$) at low density. The
combined-IV first-stage $F$ is \densHighF{}, \firstStageFBoth{}, and
\densLowF{} respectively; identification is strongest under the
low-density schedule.

\input{../results/tables/table_15_density_schedules.tex}

\subsection{Sectoral Activity}
\label{sec:sectoral}

Table~\ref{tab:sectoral_iv} reports estimates of
equation~\eqref{eq:main} for five sectoral outcomes drawn from the
industrial and agricultural censuses. Panel~A covers three
manufacturing outcomes from the 1954 and 1985 industrial censuses:
the log change in the number of establishments, in the value of
production, and in the total wage mass. Panel~B covers two
agricultural outcomes from the 1960 and 1988 agricultural censuses:
the log change in the number of farms and in the total farmed area.
Column structure matches Table~\ref{tab:population_iv}.

For manufacturing establishments, the OLS estimate is \estabOLSCoef{} (\estabOLSSE{})
and the combined IV estimate is \estabIVBCoef{} (\estabIVBSE{}). Neither is
statistically distinguishable from zero.

For manufacturing production value, the OLS estimate is \valprodOLSCoef{}
(\valprodOLSSE{}), significant at the five-percent level. The IV-LP estimate is
\valprodIVLPCoef{} (\valprodIVLPSE{}), significant at the
five-percent level. The combined IV estimate is \mfgValprodCoefIVBoth{}
(\mfgValprodSEIVBoth{}),
significant at the one-percent level. The sample is \valprodN{} districts.

For manufacturing wage mass, the OLS estimate is \massalOLSCoef{} (\massalOLSSE{}),
significant at the one-percent level. The IV-LP estimate is \massalIVLPCoef{}
(\massalIVLPSE{}), significant at the five-percent level. The combined IV
estimate is \mfgWageMassCoefIVBoth{} (\mfgWageMassSEIVBoth{}),
significant at the one-percent level.
The sample is \massalN{} districts.

For the number of agricultural farms (Panel~B), the OLS estimate is
$\farmsOLSCoef{}$ (\farmsOLSSE{}). The combined IV estimate is \farmsIVBCoef{} (\farmsIVBSE{}).
Neither is statistically distinguishable from zero.

For total agricultural farmed area, the OLS estimate is $\areaOLSCoef{}$
(\areaOLSSE{}). The combined IV estimate is \areaIVBCoef{} (\areaIVBSE{}). Neither is
statistically distinguishable from zero.

The estimates differ substantially across the two panels. The
manufacturing outcomes related to activity intensity --- production
value and wage mass --- respond strongly and positively to
$\Delta \ln \mathrm{MA}^{\mathrm{full}}_i$: combined-IV point estimates
are \mfgValprodCoefIVBoth{} and \mfgWageMassCoefIVBoth{}, respectively. The number of manufacturing
establishments does not respond detectably. The agricultural
outcomes show point estimates close to zero with wide confidence
intervals. Section~\ref{sec:mechanisms} returns to the question of
whether this pattern is consistent with the scale-economies
hypothesis introduced in Section~\ref{sec:history}.

\paragraph{Sector-matched cost schedules.}
The estimates above use the medium-density cost schedule for all
outcomes. The \citet{baumgartnerpalazzo1969} schedules assign each
sector a cargo profile: bulk agricultural freight travels at high
density, where rail is the cheap mode, and small-batch manufacturing
freight at low density, where road is
(Table~\ref{tab:cost_params}). Table~\ref{tab:sector_matched}
re-estimates each sectoral outcome with the market-access measure
constructed under its matched schedule, switching the treatment, both
instruments, and the baseline log-MA control together. The
manufacturing estimates strengthen under their matched
(road-favouring) measure: production value rises from
\mfgValprodCoefIVBoth{} to \smValprodIVBCoef{} (\smValprodIVBSE{}) and
wage mass from \mfgWageMassCoefIVBoth{} to \smMassalIVBCoef{}
(\smMassalIVBSE{}), with combined-IV first-stage $F$ of \smMfgF{} and
\smMassalF{} respectively; the number of establishments remains
statistically indistinguishable from zero, as at the baseline
schedule. The
agricultural estimates remain statistically indistinguishable from
zero under their matched (rail-favouring) measure (farms
\smNexpIVBCoef{}, farmed area \smAreaIVBCoef{}; $F = \smAgrF{}$). The
sectoral contrast of the previous paragraph is therefore not an
artifact of evaluating both sectors at the general-purpose cost
schedule.

\input{../results/tables/table_16_sector_matched.tex}

Three facts connect this sectoral pattern to the cost structure
described in Section~\ref{ssec:scale_economies}. First, the
\citet{baumgartnerpalazzo1969} schedule makes the modal shift
asymmetric across sectors by construction: moving freight from rail
to road lowers unit costs at low cargo density (8.266 versus 13.490
1960~pesos per ton-km) and raises them at high density (1.000 versus
0.465), so a restructuring that replaced rail with road capacity
changed manufacturing's transport costs in the opposite direction
from agriculture's. Second, the outcomes align with that asymmetry:
the sector whose characteristic shipments became cheaper to move
responds strongly (Table~\ref{tab:sectoral_iv}), and the response
strengthens when its market access is measured under its own cost
schedule (Table~\ref{tab:sector_matched}), while the sector whose
shipments became more expensive shows no response under either
measurement. Third, the same gradient appears for the aggregate
outcome: the population elasticity rises monotonically as the cost
schedule shifts from rail-favouring to road-favouring
(Table~\ref{tab:density_schedules}). The interpretive weighing of
these facts --- what they imply about scale economies and the modal
shift as industrial policy --- is deferred to
Section~\ref{sec:sector_mode}.

The manufacturing IV--OLS gaps run in the same direction as the
population gap of Section~\ref{sec:population} and are larger in
proportion; the two mechanisms discussed there --- selection of
closures into declining districts and measurement error in the
constructed regressor --- apply to the sectoral outcomes as well,
and we do not apportion between them here either.
% NOTE for Cote: a wage-mass-specific composition story (wage levels
% rising in declining districts as worker composition changed) was
% floated at drafting; not included pending discussion.

\input{../results/tables/table_10_sectoral_iv.tex}

\subsection{Other Outcomes}
\label{sec:other_outcomes}

Table~\ref{tab:other_outcomes_iv} reports estimates of
equation~\eqref{eq:main} for four outcomes drawn from IPUMS
microdata: the level change in the share of adults with college
education, the level change in the share with secondary education,
the level change in the share of residents who migrated in the past
five years, and the level change in the employment rate (the share
of individuals classified as employed in IPUMS EMPSTAT). All four
outcomes are level differences in population shares between 1970 and
1991. The regressor remains $\Delta \ln \mathrm{MA}^{\mathrm{full}}_i$
computed over 1960--1986; infrastructure restructuring predates the
1970 outcome measurement. Column structure matches
Table~\ref{tab:population_iv}.

For the college share, all four estimates are statistically
indistinguishable from zero, with point estimates between $\collegeMin{}$
and $\collegeMax{}$. For the secondary share, the OLS and combined-IV
point estimates are \secondaryOLSCoef{} and \secondaryIVBCoef{}; the IV-LP estimate alone is \secondaryIVLPCoef{}
($p=\secondaryIVLPP{}$).

For the recent-migration share, the OLS estimate is $\migrationOLSCoef{}$ (\migrationOLSSE{}),
significant at the one-percent level. The combined IV estimate is
\migrationCoefIVBoth{} (\migrationSEIVBoth{}), significant at the
one-percent level. The sample
is \mainPopNIVBoth{} districts. A one-unit increase in $\Delta \ln
\mathrm{MA}^{\mathrm{full}}_i$ corresponds to a \migrationCoefAbs{} decrease in the
share of residents reporting recent migration.

For the employment rate, the OLS estimate is $\employmentOLSCoef{}$ (\employmentOLSSE{}) and the
combined IV estimate is $\employmentIVBCoef{}$ (\employmentIVBSE{}), significant at
the five-percent level.

The direction of the recent-migration coefficient runs opposite to a
naive prediction that market-access gains attract in-migrants. Two
readings are consistent with the sign. First, high-market-access
districts may have retained their existing populations, producing
lower recent-migrant shares because turnover was low. Second, districts
that lost population through out-migration to high-market-access
receiving districts may have ended with small remaining populations
among whom a given absolute inflow of recent migrants generates a
higher share. Distinguishing these readings requires data on
migration flows by origin and destination, which are beyond the
cross-sectional sample available here.

The window for these four outcomes is 1970--1991 because Argentine
IPUMS microdata begins in 1970, one decade after the start of the
estimation window for Table~\ref{tab:population_iv}. The infrastructure
restructuring and the bulk of rail closures occurred before 1970, so
the timing of outcome measurement post-dates treatment.

\input{../results/tables/table_11_other_outcomes_iv.tex}

\subsection{Robustness}
\label{sec:results_robustness}

Table~\ref{tab:robustness} reports three stress tests of the main
population elasticity estimated in
Section~\ref{sec:population} (Panel~A of
Table~\ref{tab:population_iv}): the coefficient on $\Delta \ln
\mathrm{MA}^{\mathrm{full}}_i$ is \mainPopCoefIVBoth{}
(\mainPopSEIVBoth{}) under the combined
IV specification on the full 311-district sample.

\paragraph{Panel A --- alternative trade elasticity.}
The main specification sets the trade elasticity $\theta = \thetaLow{}$.
Panel~A reports the same four specifications with $\theta = \thetaHigh{}$,
which corresponds to a higher-elasticity calibration in the
market-access literature. All MA variables, including the
treatment, the two instruments, and the baseline log-MA control,
switch to the $\theta = \thetaHigh{}$ variant. The combined IV coefficient
is \thetaHighIVBCoef{} (\thetaHighIVBSE{}). The coefficient is approximately half the
main-specification value. The direction of the effect is unchanged,
and no coefficient flips sign.

Table~\ref{tab:theta_sweep} extends the two-value comparison to the
full sweep $\theta \in [1, 12]$, recomputing market access, both
instruments, and the baseline control at each value. The combined-IV
coefficient declines monotonically in $\theta$, from
\sweepBetaThetaOne{} at $\theta = 1$ to \sweepBetaThetaTwelve{} at
$\theta = 12$, while the first-stage $F$ stays between
\sweepFMin{} and \sweepFMax{} across the grid. The level of the population elasticity is
therefore calibration-dependent; its sign is not.
Section~\ref{sec:discussion} discusses the implications for
comparisons with estimates in the literature.

\input{../results/tables/diagnostic_theta_sweep.tex}

\paragraph{Transshipment bound.}
The cost surface underlying all estimates allows free mode switching:
a least-cost path can move between rail, road, and navigation at any
grid cell at no cost (Section~\ref{ssec:ma}). The opposite extreme is
infinite transshipment: each origin--destination trip uses a single
mode plus off-network land, with the trip cost taken as the minimum
across the three single-mode cost matrices. Under this bound, the
share of districts gaining market access between 1960 and 1986 falls
from \uniGainShareBase\% to \uniGainShareUni\%, and the median
origin--destination cost ratio (1986 relative to 1960) moves from
\uniMedRatioBase{} to \uniMedRatioUni{}: the bound reshapes the
market-access distribution, but the broad
cost decline persists when no trip may switch modes, and the median
pair still becomes substantially cheaper. The population
elasticity is essentially unchanged at the bound: an OLS regression
of the population change on the unimodal market-access change yields
\uniOLSUni{} (\uniOLSUniSE{}), against \uniOLSBase{} (\uniOLSBaseSE{})
for the baseline measure on the same specification and
sample ($N = \uniN{}$).\footnote{The bound is evaluated by OLS only:
instrumenting would require unimodal versions of both instruments,
and the comparison specification omits the baseline log-MA control on
both sides (no unimodal analogue exists), so the baseline OLS
coefficient here differs from Table~\ref{tab:population_iv}
column~(1).} The truth --- finite transshipment costs at
stations and ports --- lies between the two extremes, which agree on
the population elasticity; on this evidence the free-switching
assumption is not driving that estimate.

\paragraph{Panel B --- alternative hypothetical-road instruments.}
The main specification uses the least-cost-path minimum spanning
tree as the hypothetical-road instrument. Panel~B reports results
using three alternative hypothetical networks: the Euclidean
minimum spanning tree, the full bilateral least-cost path network,
and the full bilateral Euclidean network. Holding $\theta$ at \thetaLow{},
the combined-IV coefficient on $\Delta \ln \mathrm{MA}^{\mathrm{full}}_i$
ranges from \hypoAltIVBMin{} to \hypoAltIVBMax{} across the four choices. Individual
IV-Hypo columns show wide variation; the large point estimates and
standard errors in that column reflect weak-instrument noise in
some of the alternative networks and are not interpretable in
isolation. The combined-IV column is more stable.

\paragraph{Panel C --- sample robustness.}
Panel~C refits the main specification on the \placeboNIVBoth{}-district subsample
for which the 1947 placebo outcome is defined, as reported in
Section~\ref{sec:pre_trends}. On this subsample, the OLS coefficient
is \subOLSCoef{} (\subOLSSE{}), significant at the five-percent level, and the
combined IV coefficient is \subIVBCoef{} (\subIVBSE{}), significant at the
five-percent level. Both estimates are larger than their
full-sample counterparts.

The Panel~C estimate bears on the interpretation of
Section~\ref{sec:pre_trends}. The placebo coefficient on the
1947--1960 population change was \placeboCoefIVBoth{} on the same
\placeboNIVBoth{}-district
subsample. The main-specification coefficient on the 1960--1991
population change on the same subsample is \subIVBCoef{}. The pre-reform
coefficient and the post-reform coefficient are of the same order
of magnitude on the common sample. This co-movement is consistent
with both candidate readings offered in Section~\ref{sec:pre_trends}
and does not adjudicate between them.

\input{../results/tables/table_12_robustness.tex}
