% ===========================================================================
% Section 1: Introduction
%
% Ported from paper/paper_planned.pdf (the authoritative coauthor draft).
% Follows Shapiro's contractual-intro template (see Plan/foursteps.pdf):
%   ¶1-2 Motivation        ¶3-4 Challenges         ¶5 This paper
%   ¶6-7 Model/framework   ¶8-9 Data               ¶10-11 Methods
%   ¶12-13 Findings        ¶14-15 Contributions    ¶16-17 Roadmap
%
% CONTRACTUAL PASS (Shapiro step 4, provisional-draft assembly):
%   - All three findings paragraphs filled (main, counterfactuals,
%     mechanisms); bracketed numbers resolved against Section 2's
%     verified figures; [312] -> \nDistricts macro.
%   - Broken promises repaired: "sectoral employment" -> "sectoral
%     economic activity" (Table 10 measures activity, not employment);
%     "spatial placebo tests, and over-identification tests" removed
%     (neither is reported in the paper; see matching fix in Section
%     4.6). The validation sentence now promises exactly what Sections
%     4-5 deliver, including owning the not-clean pre-trends.
% ===========================================================================

\section{Introduction}
\label{sec:intro}

How do large-scale transport infrastructure changes reshape regional
economies? While most research examines infrastructure expansions
designed to promote development, governments often face a different
challenge: fiscal consolidation that forces difficult choices about
which infrastructure to maintain. Argentina's experience between 1960
and 1991 provides a unique window into this question. Facing severe
fiscal pressure, the government implemented a massive transport
restructuring: roughly 10{,}000 kilometers of railway lines were
closed, while the paved road network expanded from roughly 10{,}000
to 28{,}000 kilometers between 1954 and 1986. This episode offers
rare evidence on how fiscal consolidation-driven infrastructure
changes—rather than development-oriented investments—affect regional
market access and economic outcomes.

This setting is valuable for two reasons. First, it speaks to a
policy-relevant question: when governments must cut infrastructure
spending, how do these cuts reshape the economic geography? The
Argentine plan, motivated by fiscal rationalization rather than
regional development goals, effectively acted as an unintended
industrial and regional policy by shifting the dominant transport
mode from rail to road. Second, it provides an opportunity to estimate
how regional outcomes respond to market access changes in a multimodal
infrastructure shock. Unlike typical infrastructure studies that
examine single-mode expansions, Argentina's bundled rail-road
restructuring allows us to study elasticities of population and
sectoral economic activity to market access changes that integrate
both transport modes.

Estimating the causal effects of such infrastructure changes faces
two fundamental challenges. First, infrastructure placement and
removal decisions are endogenous. Even when motivated by fiscal
consolidation, governments may close lines in already-declining
regions or expand roads where growth is expected. This makes it
difficult to separate the causal effect of market access changes
from pre-existing economic trends.

Second, when both rail and road networks change simultaneously, their
effects are difficult to disentangle. The changes are often spatially
correlated—regions losing rail access may or may not gain road
access—creating multicollinearity problems if we try to estimate
mode-specific effects in a single regression. Moreover, the
theoretical framework suggests that what matters for regional
outcomes is total market access integrating all transport modes, not
the separate contribution of each mode. This raises a methodological
question: what can and cannot be identified about mode-specific
effects in multimodal infrastructure episodes?

This paper addresses these challenges by studying Argentina's
transport restructuring through the lens of market access. We
construct a unified market access index that integrates both rail and
road networks using a standard spatial economics framework, computing
generalized transport costs and shortest-path routes on the historical
networks. Our main analysis estimates how regional outcomes—population,
sectoral economic activity, and additional outcomes including
education and migration—respond to changes in
total market access induced by the observed infrastructure plan. We
then use counterfactual market access measures to explore the relative
roles of rail closures versus road expansion, constructing separate
scenarios where only rail changes (roads frozen) or only roads change
(rail frozen).

Our empirical framework follows the market access approach standard
in spatial economics and trade. We model each district's economic
outcomes as functions of its market access, defined as the
population-weighted sum of inverse generalized transport costs to all
other districts. To address endogeneity, we instrument for observed
market access changes using two sources of variation. First, we
exploit the Larkin Plan's technical rules: the World Bank study that
motivated the reforms only examined railway segments below a specific
freight threshold for potential closure, creating a discontinuity in
closure probability. Second, we construct hypothetical road networks
connecting major cities via least-cost paths, which predict road
expansion but are orthogonal to local economic shocks.

We construct a novel dataset combining historical transport networks
with census data at the district level, covering \nDistricts{}
districts across Argentina. We validate our identification strategy
with pre-period balance checks and a pre-trends placebo, examine
sensitivity to spatially correlated standard errors in
Section~\ref{sec:discussion}, and report where the validation
package falls short: the pre-trends placebo rejects a clean null at
the five-percent level in the OLS and combined-IV specifications, a
rejection that spatially robust standard
errors do not soften, and we state that limitation
rather than obscure it.

% Findings paragraphs (counterfactuals, mechanisms) filled from the
% drafted Sections 6-7; framing per Plan/provisional_decisions_log.md
% F1 and F2. NOTE: the original placeholder expected "rail-driven MA
% changes account for most effects" -- the data said otherwise (similar
% point estimates, rail better identified), and the paragraph states
% what the data show.

The counterfactual decomposition qualifies the modal question that
motivates the paper. Recomputing market access with one network frozen
at its pre-reform configuration, we find that the only-rail and
only-road changes contribute similar point estimates to the population
response (\cfRailCoefIVPop{} and \cfRoadCoefIVPop{}, against
\mainPopCoefIVBoth{} for the full change), while
differing sharply in identification: the rail component, instrumented
by the Larkin Plan alone, is the best-identified object in the paper,
with a first-stage $F$ above 100 for total population. We find no
support
for the view that one mode's market-access channel dominates the
other; the evidence is instead consistent with population responding
to market access itself, whichever mode produces it.

The mechanism decomposition shows where that response lives. Adding
own-district infrastructure changes—rail and road kilometers and
two binary margins—to the main specification removes roughly half
of the market-access coefficient (\mainPopCoefIVBoth{} to
\mechAllZCoef{}), so about half of
the estimated response operates through infrastructure located in the
district itself and the remainder through regional connectivity
beyond it. Suggestive heterogeneity patterns point in the same
direction as theory: the market-access coefficient is larger in
initially smaller, more rural, and more remote districts, although
these interactions are too imprecisely identified to support
magnitude claims.

Our main results document a contrast between population and sectoral
activity. The elasticity of total population with respect to
$\Delta \ln \mathrm{MA}^{\mathrm{full}}$ is \mainPopCoefIVBoth{}
(\mainPopSEIVBoth{}) in the combined-IV specification, significant at
the ten-percent but not the five-percent level; the corresponding OLS
estimate is significant at the five-percent level, and the urban and
rural coefficients are positive but less precisely estimated. The
evidence for an aggregate population response is suggestive rather
than conclusive. In
contrast, the elasticity of manufacturing activity is sizable and
significant: the log change in the value of manufacturing production
has an elasticity of \mfgValprodCoefIVBoth{} (\mfgValprodSEIVBoth{})
and the log change in the manufacturing wage mass has an elasticity
of \mfgWageMassCoefIVBoth{} (\mfgWageMassSEIVBoth{}). Agricultural
activity, measured by the number of farms and by total farmed area,
does not respond detectably. The contrast survives measuring each
sector's market access under a cost schedule matched to its cargo
profile, and strengthens for manufacturing. This pattern is consistent with the
scale-economies hypothesis introduced in Section~\ref{sec:history}:
the transport restructuring shifted the cost structure faced by
different sectors differently, with manufacturing benefiting more
than agriculture. Robustness to alternative trade elasticities,
alternative hypothetical-road instruments, and subsample choice is
reported in Section~\ref{sec:results_robustness}; the level of the
population elasticity, though not the sectoral contrast, depends on
the trade-elasticity calibration, a dependence
Section~\ref{sec:discussion} quantifies.

This paper contributes to several literatures. First, we provide
causal evidence on the regional effects of infrastructure
disinvestment driven by fiscal consolidation. The Argentine episode
involved not just the removal of rail infrastructure but a
technological shift in the dominant transport mode—from rail to
road—that effectively acted as an unintentional industrial policy,
reshaping the country's economic geography. Second, we contribute to
understanding mode-specific effects in multimodal infrastructure
episodes, proposing a counterfactual market access approach that
clarifies what can and cannot be identified. Third, our findings
speak to contemporary policy debates about modal shifts in transport,
as many countries face ongoing transitions between rail and road.
Fourth, we contribute to a longstanding debate in Argentine economic
history about whether railroad closures caused the decline of
hinterland towns or merely followed pre-existing trends.

The paper proceeds as follows. Section~\ref{sec:history} provides
historical context. Section~\ref{sec:data} describes data and market
access construction. Section~\ref{sec:identification} presents the
empirical strategy. Section~\ref{sec:results} reports main results.
Section~\ref{sec:counterfactuals} explores mode-specific patterns.
Section~\ref{sec:mechanisms} examines mechanisms and heterogeneity.
Section~\ref{sec:discussion} interprets the results and states their
limitations, and Section~\ref{sec:conclusion} concludes.
